---
title: "Exploratory analysis of Nursing Homes Quality and Utilization in the US"
author: "Roland Abi"
format: html
editor: visual
execute: 
  echo: false
  warning: false
---

```{r, echo=FALSE}

library(tidyverse)
library(janitor)
library(viridis)
library(DataExplorer)
library(ggthemes)
library(gt)
library(GGally)
library(showtext)

source("code/data_cleaning.R")
source("code/exploratory_analysis.R")
```

## Introduction

Nursing homes, also known as skilled nursing facilities (SNFs) or long-term care facilities, are residential facilities that provide comprehensive healthcare and assistance to individuals who require significant support with daily activities and medical care. They primarily cater to older adults with chronic illnesses, disabilities, or those recovering from surgeries or hospital stays. Research shows that there are about 26,514 nursing homes in the US and an estimated 70% of people who reach the age of 65 will need long term care at some point in life.

## About the data

This dataset provides comprehensive information on currently active nursing homes, such as the type of ownership, number of certified beds, legal business name, and continuing care retirement community status. It also includes ratings and scores associated with quality measures, staffing levels and other factors used in the Five-Star Rating System. Data are presented as one row per nursing home to give an overview of each facility's characteristics. This dataset is a valuable resource for healthcare professionals and industry professionals who need access to detailed information on the operations of nursing homes.

## Methods

The Project utilized the tidyverse, ggthemes and the viridis package to develop plots and summary tables.

## Exploring dataset

```{r}
plot_missing(
  df4,
  title = str_glue(
    'Exploring missing data (N = {count(df4)})'
  )
)

```

## Highest percentage of nursing homes

```{r}
#| fig-cap: |
#|  Top 10  states with the highest percentage of nursing homes
#| label: fig-top-10-states
#| output-location: slide
plt2
```

## Nursing home utilization

```{r}
#| label: fig-nursing-home-utilization
#| fig-cap: |
#|  Nursing homes with >= 500 bed capacity and their utilization

plt3
```

## Overall rating of nursing homes

```{r}
#| label: fig-overall-rating
#| fig-cap: |
#|  Overall rating of nursing homes with over 700 bed space
#| fig-format: svg

plt6
```

## Overall staff rating and incidents

```{r}
#| label: tbl-summary-of-each-nursing-home
#| tbl-cap: Description of overall staff rating, beds and incidents
tbl1
```

## Total health score

```{r}
#| label: fig-total-health-score
#| fig-cap: |
#|  Total health score between cycle 1 and cycle 2
#| fig-align: center
#| fig-format: pdf
#| fig-width: 10
#| fig-height: 4
plt7
```

## Average staff hours

```{r}
#| label: tbl-average-staff-hours
#| tbl-cap: Average staff hours across states
tbl2
```

## Average health revisit score

```{r}
#| label: fig-health-revisit-score
#| fig-cap: Average health revisit score
plt8
```

## Note 1:

@fig-top-10-states shows the top states with the highest percentage of nursing homes. The analysis showed the state of California (7.74%) as having the highest percentage of nursing home, closely followed by Texas (7.72%). @fig-nursing-home-utilization shows the percentage utilization of nursing homes with 500 beds or more and their ownership type. We observed that majority of government owned nursing homes experience on average above 90% utilization with the exception of Henry J Carter Skilled Nursing Facility.

## Note 2:

@fig-overall-rating explored the overall rating of nursing homes with over 700 bed space in US. Laguna Honda Hospital and Rehabilitation Center in California had an overall rating of 4 out of 5; Iowa Veterans Home in Iowa, The New Jewish Home, Harry and Jeanette Weinberg, Kings Harbor Multicare Center and Henry J Carter Skilled Nursing Facility In New York respectively, had an overall rating 5. Meanwhile Coler Rehabilitation and Nursing Care Center and Isabella Geriatric Center Inc also in New York, had an overall rating of 4 respectively. Furthermore, Fair Acres Geriatric Center and Glendale Northampton County in Pennsylvania had an overall rating of 1 and 4 respectively.

## Note 3:

@tbl-summary-of-each-nursing-home displays the overall staff rating number of residents in certified beds, facility reported incidents and number of fines. Spectrum Health Rehab & Nursing Center-Fuller Ave with an overall staff rating of 5 had the highest reported facility incidents as well as the highest number of fines, followed by Arden House with 20 reported incidents and 3 fines. The analysis showed that facilities with 200 or more residents in certified beds have experienced at least one incident and at least one fine.

## Note 4:

Total health score attained at cycle 1 and cycle 2 for nursing homes with 700 or more bed spaces was also explored. Fair Acres Geriatric Center gained considerable increase in total health score from 72 in cycle one, to 114 in cycle 2. Laguna Honda Hospital & Rehabilitation Center, experienced some increase in total health score although not as pronounced as Fair Acres Geriatric Center. Other nursing home that experienced a similar trend were Isabella Geriatric Center Inc and Iowa Veterans Home (@fig-total-health-score )

## Note 5:

@tbl-average-staff-hours shows the top 10 states with the highest average staffing hours. Alaska has had the highest CNA staffing hours on average, while in the District of Columbia, Florida and Hawaii average working hours for CNA staff was 2.8hours. Average working hours for LPN staffs was between 0.4 and 1.0 hours. Alaska also recorded the highest average (2.1) working hours for RN staff.

## Note 6:

Average health revisit score across three cycles was explored. It was observed that Apex Center and Barnes-Kasson County Hospital which offered Medicare and Medicaid services in Idaho and Pennsylvannia reported a somewhat similar revisit score across the three cycles. Health revisit score for Carney Nursing Center located in Kansas which had provision for only Medicaid showed a high-low-high trend across the three cycles. Lawrence Hall Nursing Center located in Arizona reported a high revisit score in cycle 2 both was lower in cycle 1 and 3 (@fig-health-revisit-score )

## Credits

-   The Datavastor

-   https://www.kaggle.com/datasets/thedevastator/nursing-home-quality-staffing

-   https://data.world/health
